Analysis plan

### 2023-03-08 02. Design/MSG-P2312-Analysis Plan-1-clusters.rmd
Update MSG-P2312-Analysis Plan-1-clusters.rmd
Updating based on Mike's feed back, creating tables of variables from the data, added more detail to the cohort, explaining joins and deduplication in more detail, updating missing data table to reflect combined data, removed rule of thumb note for cluster number

@@

```{r}
tbl <- tibble("Variable Name" = c(
  "date",
"SSO",
"short_id",
"dept_id",
"employee_name",
"supervisor_name",
"location",
"daily_status",
"supervisor_correction"
),
"Description" = c(
  "Date of check-in",
  # Is SSO single sign on? What did we decide?
"",
"2-digit department id",
"Full department id code",
"Employee name (Last,First Middle)",
"Supervisor name (Last,First Middle)",
"State abbreviation-City-Site (e.g., home or address); could string split and match state with roster, possibly city too.",
"E.g., Teleworking, Leave, Supervisor Corrected, No Response, Reporting to Job Site, Other",
# I think that this is an indicator for whether the supervisor corrected the record
""
)
)
```

@@

```{r}
tbl <- tibble("Variable Name" = c(
"Snapshot.Date",
"Department.ID",
"Employee.Job.Business.Unit.Description",
"Full.Name",
"Employee.E.Mail.Address",
"Position.Title",
"Pay.Plan",
"Occupational.Series",
"Grade",
"EOD.Date",
"Supervisor.Level",
"Supervisor.Level.Description",
"Supervisor.Display.Name",
"Supervisor.E.Mail.Address",
"Business.Line.2.Letter.Department.ID.",
"GSA.SSO",
"Agency.Subelement.Code.Description",
"GSA.Region",
"GSA.Duty.Station.Region",
"Location.City",
"Location.State",
"Duty.Station.Code",
"Employee.Job.Bargaining.Unit.Description",
"Employee.Summary.Bargaining.Unit",
"Payroll.Accounting.Code",
"Position.Number",
"Position.Category.Code",
"Position.Category.Description"
),
"Description" = c(
 "Date of data export",
"Should align with check-in’s 'dept_id'",
"",
"Last,First Middle; should align with check-in’s 'employee_name'",
"",
"",
"",
"",
"",
"",
"",
"",
"First, Last",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"",
"E.g., Onsite flexible, Offsite, Onsite required, Not specified, Position has changed"
)
)
```

@@

For all analysis, we will used a defined cohort of staff to examine work
status over period of time. A cohort of staff employed in the month of
March 2022 will be followed over six months. Teams will be defined by
the supervisors of groups of staff in the beginning of March 2022. Any
analysis we conduct may be reproduced on a separate cohort or may be
combined with multiple cohorts depending on the data availability.

Both data sources will be checked for duplication and deduplicated. The
unique identifiers of the daily check-in data are `employee_name` and
`date`. For the roster data the unique identifier is `Full.Name` and
`Snapshot.Date`. The general approach to deduplication will be that
categorical variables will be restructured as indicator variables taking
on a one for each category associated with the duplicate. From the
roster data, if the duplication is based off of `EOD.Date`, the later
date will be selected along with the other variables associated with
that date. There appear to be no numeric variables.

@@

<!-- [I might be more discerning about using "staff" vs. "employees" vs. "teams" vs. "groups" vs. "clusters" -- we are grouping employees into teams, but also categorizing them.]  -->
We will approach grouping staff and teams in two ways. First,
rules-based groups will be formed. For these, definitions of different
types of work will be defined. Second, groups will be formed using
k-means clustering. This unsupervised machine learning method produces a
defined number of groups (or clusters) minimizing the distance between
the members.

@@

Missing data from daily check in will be treated as defined in the table below.

@@

```{r}
tbl <- tibble("Variable Name" = c(
  "date",
"SSO",
"short_id",
"dept_id",
"employee_name",
"supervisor_name",
"location",
"daily_status",
"supervisor_correction"
),
"Treatment if missing" = c(
  "Drop observation",
  "",
  "Impute from other observations of same staff",
  "Impute from other observations of same staff",
  "Drop observation",
  "Impute from other observations of same staff",
  "Label as missing",
  "Label as missing",
  "Label as missing"
)
)
```

Missing data from the roster will be treated as show in the table below.

#### Table `r table_n`. Treatment of missing data from roster

```{r}
tbl <- tibble("Variable Name" = c(
"Snapshot.Date",
"Department.ID",
"Employee.Job.Business.Unit.Description",
"Full.Name",
"Employee.E.Mail.Address",
"Position.Title",
"Pay.Plan",
"Occupational.Series",
"Grade",
"EOD.Date",
"Supervisor.Level",
"Supervisor.Level.Description",
"Supervisor.Display.Name",
"Supervisor.E.Mail.Address",
"Business.Line.2.Letter.Department.ID.",
"GSA.SSO",
"Agency.Subelement.Code.Description",
"GSA.Region",
"GSA.Duty.Station.Region",
"Location.City",
"Location.State",
"Duty.Station.Code",
"Employee.Job.Bargaining.Unit.Description",
"Employee.Summary.Bargaining.Unit",
"Payroll.Accounting.Code",
"Position.Number",
"Position.Category.Code",
"Position.Category.Description"
),
"Treatment if missing" = c(
  "Label as missing",
  "Label as missing",
  "Label as missing",
  "Drop observation",
  "Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing"
)
)
```

@@

To determine the number appropriate number of clusters, several methods
exist. A rule of thumb to identify the correct number is to use the
number calculated from:

$$
\frac{\sqrt{NumberOfObservations}}{2}
$$

However, for this work, the *elbow method* will be used. To do this, the
k-means algorithm will be executed for different numbers of clusters
from 1 to 20. For each number of clusters, a total within sum of squares
will be calculated. As the number of number of clusters increases the
total within sum of squares will decrease. The optimal number of
clusters is when the decrease in total within sum of squares starts
tapering off.

>>>>

<!-- # MIKES COMMENTS -->

<!-- Proofread for typos -->

<!-- The analysis plan is a good place to define the rules for classification -->
<!-- Re: missing supervisor name and some other values. It is possible that -->
<!-- people change supervisors or departments. Do we care about this? And how -->
<!-- should we deal with it? At a minimum, maybe we want a slightly different -->
<!-- imputation like if supervisor is the same both with earlier and later -->
<!-- nonmissing observations, use that value, otherwise use prior supervisor? -->

<!-- How are you doing team classification? Are you going to generate a team -->
<!-- ID based on some rule? If so what is it, and to the prior point, how do -->
<!-- you handle changing team composition over time? -->

<!-- Not sure about using percentiles for cutoffs. I would assume based on -->
<!-- classifications that more that 5 percent of teams are fully remote. For -->
<!-- individuals, i think there is more alignment on things like number of -->
<!-- days in a pay period in the office because of policy definitions for -->
<!-- people who are not remote and have to be in the office 2 days per pay -->
<!-- period. For teams maybe an extension of that basic ideas (e.g., fewer -->
<!-- than 2 days in office on average is fully remote) -->

<!-- Are all of these taking averages over the whole time period (9 months or -->
<!-- whatever)? Would we potentially want to know about things like people -->
<!-- who have more specific weekly patterns? -->

<!-- The thing that is missing here that is potentially more complicated is -->
<!-- figuring out more about how teams operate as collections of individuals. -->
<!-- For example, you could I think in this approach have a team that is 20 -->
<!-- percent average telework where everyone goes in person on Mondays and -->
<!-- the rest of the time is distributed, or you could have a team that is 20 -->
<!-- percent telework where people go in at random times and never see each -->
<!-- other. I am interested to know the extent to which people are -->
<!-- co-locating with their teams (and if some people are being left -->
<!-- out---which may help motivate some equity questions about performance -->
<!-- ratings for people who don't get face time with the boss, for example). -->
<!-- I think we are missing that level of analysis here. -->

@@

<!-- Table 1 and 2 could use better updating of description column -->

<!-- SSO=Staff or service office -->

<!-- Done -->

```{r}
tbl_daily_check_in <- tibble("Variable Name" = c(
  "date",
"sso",
"short_id",
"dept_id",
"employee_name",
"supervisor_name",
"location",
"daily_status",
"supervisor_correction"
)
)
```

```{r}
tbl <- 
  tbl_daily_check_in %>% 
  add_column(
    "Description" = c(
  "Date of check-in",
  # Is SSO single sign on? What did we decide?
"Staff or service office",
"2-digit department id",
"Full department id code",
"Employee name (Last, First Middle)",
"Supervisor name (Last, First Middle)",
"State abbreviation-City-Site (e.g. MA-WESTFORD-HOME and MA-BOSTON-FB Causeway St)",
"Status of the staff on that day (e.g., Teleworking, Leave, Supervisor Corrected, No Response, Reporting to Job Site, Other)",
# I think that this is an indicator for whether the supervisor corrected the record
"If daily status is Supervisor Corrected, then this is the work status entered by the supervisor"
)
)
```

@@

```{r}
tbl_roster <- tibble("Variable Name" = c("snapshot_date", "department_id", "employee_job_business_unit_description", "full_name", "employee_e_mail_address", "position_title", "pay_plan", "occupational_series", "grade", "eod_date", "supervisor_level", "supervisor_level_description", "supervisor_display_name", "supervisor_e_mail_address", "business_line_2_letter_department_id", "gsa_sso", "agency_subelement_code_description", "gsa_region", "gsa_duty_station_region", "location_city", "location_state", "duty_station_code", "employee_job_bargaining_unit_description", "employee_summary_bargaining_unit", "payroll_accounting_code", "position_number", "position_category_code", "position_category_description"
                                  # , "file_name"
)
)
```

```{r}
tbl <- 
  tbl_roster %>% 
  add_column(
"Description" = c(
 "Date of data export",
"GSA department ID",
"Business unit of the staff person",
"Employee name as Last,First Middle",
"Email address of staff",
"Position title",
"Pay plan code",
"Occupational series code",
"Position grade",
"Entry on duty i.e. staff's first day on the job",
"Supervisor status code of staff",
"Supervisor status description of staff",
"Supervisor name as First Last",
"Supervisor email address",
"First two letters of GSA department ID",
"Staff or service office code",
"Staff or service office description",
"GSA region number",
"Duty station GSA region number",
"Staff city",
"Staff state",
"Duty station number",
"Staff bargaining unit descripton",
"Staff bargaining unit number",
"Payroll accounting code",
"Position number",
"Position location category code",
"Position location category description (e.g., Onsite flexible, Offsite, Onsite required, Not specified)"
)
)
```

@@

<!-- Why start cohort in March? I believe because of data availability, but good to clarify in text. -->

<!-- Done -->

For all analysis, we will used a defined cohort of staff to examine work
status over period of time. A cohort of staff employed in the month of
March 2022 will be followed over six months. Teams will be defined by
the supervisors of groups of staff in the beginning of March 2022. We
are choosing March 2022 because we were told by the data providers that
earlier data was less reliable. We also wanted to have enough time to
pass to examine work patterns. March 2022 is a good start for both data
reliability and length of time to examine behavior. Any analysis we
conduct may be reproduced on a separate cohort or may be combined with
multiple cohorts depending on the data availability.

<!-- Not sure i follow the deduping logic. If two employees have the same name, then they may have different EOD Dates. But maybe there are no duplicates on full name so it’s not an issue. -->

Both data sources will be checked for duplication and deduplicated. The
daily check-in data has no true unique identifier. The roster data is
uniquely identified by `employee_e_mail_address` and `snapshot_date`.
The roster data *should not* have any duplications on
`employee_e_mail_address` and `snapshot_date`, but this will be checked.
To create a unique identifier in the daily check-in data, we will
following process:

1.  Start with check-in data from one month (call this *month 1*)

2.  Transform supervisor name to match format of the roster data

3.  Join roster data from the previous month (call this *month 0*) on
    employee name, staff or service office, department ID and supervisor
    name

4.  Check to see if any rows are missing `employee_e_mail_address` after
    the join

5.  For any rows missing `employee_e_mail_address`, join roster data
    from data from *month 1*

6.  Check to see if any rows are missing `employee_e_mail_address` after
    the join

7.  For any rows missing `employee_e_mail_address`, join roster data
    from data from *month 2*

8.  Repeat steps 5 and 6 until no new `employee_e_mail_address` entries
    are found, or no roster data is available

9.  Repeat steps 1 through 7 for all other months in observation window

10. Bind all joined data across observation window.

After this, the combined data *should* be uniquely identified by
`employee_e_mail_address` and `date`. The data set will contain all the
daily check-in and roster information across the observation window.

If any duplicates remain after the steps above, the general approach to
deduplication will be that categorical variables will be restructured as
indicator variables taking on a one for each category associated with
the duplicate. There appear to be no numeric variables.

@@

<!-- [I might be more discerning about using "staff" vs. "employees" vs. "teams" vs. "groups" vs. "clusters" -- we are grouping employees into teams, but also categorizing them.]  -->

We will approach grouping staff and teams in two ways. First,
rules-based groups will be formed. For these, definitions of different
types of work will be defined. Second, groups will be formed using
k-means clustering. This unsupervised machine learning method produces a
defined number of groups (or clusters) minimizing the distance between
the members.

@@

Missing data from the combined data in will be treated as defined in the
table below.

@@

```{r}
tbl_combined <- 
  tbl_daily_check_in %>% 
  bind_rows(tbl_roster)
# tbl_combined
```

```{r}
tbl <- 
tbl_combined %>% 
add_column(
"Treatment if missing" = c(
  "Drop observation",
  "Joined with gsa_sso from roster data, impute from other observations of same staff",
  "Impute from other observations of same staff",
  "Joined with department_id from roster data, impute from other observations of same staff",
  "Joined with full_name from roster data, drop observation",
  "Joined with supervisor_display_name from roster data, impute from other observations of same staff",
  "Label as missing",
  "Label as missing",
  "Label as missing", "Label as missing",
  "Label as missing",
  "Label as missing",
  "Drop observation",
  "Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing",
"Label as missing"
)
) %>% 
  filter(!(`Variable Name` %in% c("gsa_sso","department_id","full_name","supervisor_display_name")))
```

@@

<!-- Why mention sq(num obs)/2 if you are not going to use that rule? -->

<!-- Removed -->

To determine the number appropriate number of clusters, several methods
exist. For this work, the *elbow method* will be used. To do this, the
k-means algorithm will be executed for different numbers of clusters
from 1 to 20. For each number of clusters, a total within sum of squares
will be calculated. As the number of number of clusters increases the
total within sum of squares will decrease. The optimal number of
clusters is when the decrease in total within sum of squares starts
tapering off.
